Predicting and Preventing Employee Turnover through HR Analytics

### 2024-06-15 code/eda-code/HReda.qmd
Project part 2 - second update

This Quarto file loads the cleaned data and does some exploring.

I'm only showing it the way where the code is included in the file. 
As described in the `processing_code` materials, I currently prefer the approach of having R code in a separate file and pulling it in.

But I already had this written and haven't yet re-done it that way. Feel free to redo and send a pull request on GitHub :)

Again, it is largely a matter of preference and what makes the most sense to decide if one wants to have code inside Quarto files, or as separate R files.
And sometimes, an R script with enough comments is good enough and one doesn't need a Quarto file.

Also note that while here I split cleaning and exploring, this is iterative. You saw that as part of the processing, we already had to explore the data somewhat to understand how to clean it. In general, as you explore, you'll find things that need cleaning. As you clean, you can explore more. Therefore, at times it might make more sense to combine the cleaning and exploring code parts into a single R or Quarto file. Or split things in any other logical way.

As part of the exploratory analysis, you should produce plots or tables or other summary quantities for the most interesting/important quantities in your data. Depending on the total number of variables in your dataset, explore all or some of the others. Figures produced here might be histograms or density plots, correlation plots, etc. Tables might summarize your data.

Start by exploring one variable at a time. Then continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in. If your dataset is small, you can do that for all variables.

Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones.

@@

```{r}
col_names = names(myhrdata)
num_columns <- 3

num_rows <- ceiling(length(col_names) / num_columns)

summary_matrix <- matrix(col_names, nrow = num_rows, byrow = TRUE)

summary_df <- as.data.frame(summary_matrix)

# Print the summary data frame
print(summary_df)
#Save to file
summarytable_file = here("results", "tables", "HRvariablestable.rds")
saveRDS(summary_df, file = summarytable_file)

@@

Gender and Tenure Boxplot

@@

# Data exploration through tables

Showing a bit of code to produce and save a summary table.

```{r}
summary_df = skimr::skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results", "tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

We are saving the results to the `results` folder. Depending on how many tables/figures you have, it might make sense to have separate folders for each. And/or you could have separate folders for exploratory tables/figures and for final tables/figures. Just choose a setup that makes sense for your project and works for you, and provide enough documentation that someone can understand what you are doing.

# Data exploration through figures

Histogram plots for the continuous outcomes.

Height first.

```{r}
p1 <- mydata %>% ggplot(aes(x=Height)) + geom_histogram() 
plot(p1)
figure_file = here("results", "figures","height-distribution.png")
ggsave(filename = figure_file, plot=p1) 
```

Now weights.

```{r}
p2 <- mydata %>% ggplot(aes(x=Weight)) + geom_histogram() 
plot(p2)
figure_file = here("results", "figures","weight-distribution.png")
ggsave(filename = figure_file, plot=p2) 
```

Now height as function of weight.

```{r}
p3 <- mydata %>% ggplot(aes(x=Height, y=Weight)) + geom_point() + geom_smooth(method='lm')
plot(p3)
figure_file = here("results", "figures", "height-weight.png")
ggsave(filename = figure_file, plot=p3) 
```

Once more height as function of weight, stratified by gender. Note that there is so little data, it's a bit silly. But we'll plot it anyway.

```{r}
p4 <- mydata %>% ggplot(aes(x=Height, y=Weight, color = Gender)) + geom_point() + geom_smooth(method='lm')
plot(p4)
figure_file = here("results", "figures", "height-weight-stratified.png")
ggsave(filename = figure_file, plot=p4) 
```

# Notes

For your own explorations, tables and figures can be "quick and dirty". As long as you can see what's going on, there is no need to polish them. That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.). Those should look as nice, polished and easy to understand as possible.

>>>>

@@

```{r}
col_names = names(myhrdata)
num_columns = 3

num_rows = ceiling(length(col_names) / num_columns)

summary_matrix = matrix(col_names, nrow = num_rows, byrow = TRUE)

summary_df = as.data.frame(summary_matrix)

print(summary_df)
#Save to file
summarytable_file = here("results", "tables", "HRvariablestable.rds")
saveRDS(summary_df, file = summarytable_file)

@@

Tenure Distribution

@@

```{r}
p2 = myhrdata %>%
  ggplot(aes(x = TenureCategory)) +
  geom_bar(fill = "gray", color = "black") +
  labs(x = "Tenure", y = "Count", title = "Distribution of Tenure Categories") + 
  theme_minimal() +
  theme(panel.background = element_rect(fill = "white"),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = FALSE)

print(p2)

figure_file = here("results", "figures","tenure-category.png")
ggsave(filename = figure_file, plot=p2)
```

```{r}
education_levels <- c("Below College", "College", "Bachelor's", "Master's", "Doctorate")

# Convert Education to factor with defined levels
myhrdata$Education <- factor(myhrdata$Education, levels = education_levels)

# Boxplot of MonthlyIncome by Education
p3 <- ggplot(myhrdata, aes(x = Education, y = MonthlyIncome)) +
  geom_boxplot(fill = "gray", color = "black") +
  labs(x = "Education", y = "Monthly Income", title = "Boxplot of Monthly Income by Education") + 
  theme_minimal() +
  theme(panel.background = element_rect(fill = "white"),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Print the plot
print(p3)

figure_file = here("results", "figures","income-education.png")
ggsave(filename = figure_file, plot=p3)
```

### 2024-06-15 code/processing-code/HRprocessingfile.qmd
Project part 2 - second update

---
title: "Cleaning HR Data for Analysis"
author: "Guillermo Gallardo"
date: "Summer 2024"
output: html_document
---

# Processing script

This contains the same code and comments/information as `processingcode.R`.

This just shows it as Quarto file, to give you an idea how to do it in a setup that combines code and text in a single file.

See the other Quarto file for my currently preferred approach of pulling code from the R script into the Quarto file.

@@

```

```{r}
HRprocesseddata = d2
```

@@

# Notes

Removing anyone who had "faulty" or missing data is one approach. It's often not the best. based on your question and your analysis approach, you might want to do cleaning differently (e.g. keep individuals with some missing information).

# Cleaning DELETE WHEN DONE

By inspecting the data as done above, we find some problems that need addressing:

First, there is an entry for height which says "sixty" instead of a number. 
Does that mean it should be a numeric 60? It somehow doesn't make sense since the weight is 60kg, which can't happen for a 60cm person (a baby).
Since we don't know how to fix this, we might decide to remove the person. This "sixty" entry also turned all Height entries into characters instead of numeric. That conversion to character also means that our summary function isn't very meaningful. So let's fix that first.

```{r}
d1 <- rawdata %>% dplyr::filter( Height != "sixty" ) %>% 
                  dplyr::mutate(Height = as.numeric(Height))
skimr::skim(d1)
hist(d1$Height)
```

Now we see that there is one person with a height of 6. That could be a typo, or someone mistakenly entered their height in feet. Since we unfortunately don't know, we might need to remove this person, which we'll do here.

```{r}
d2 <- d1 %>% dplyr::mutate( Height = replace(Height, Height=="6",round(6*30.48,0)) )
skimr::skim(d2)
```

Height values seem ok now.

Now let's look at the `Weight` variable. There is a person with weight of 7000, which is impossible, and one person with missing weight.
To be able to analyze the data, we'll remove those individuals as well.

```{r}
d3 <- d2 %>%  dplyr::filter(Weight != 7000) %>% tidyr::drop_na()
skimr::skim(d3)
```

Now checking the `Gender` variable. Gender should be a categorical/factor variable but is loaded as character. We can fix that with simple base R code to mix things up.

```{r}
d3$Gender <- as.factor(d3$Gender)  
skimr::skim(d3)
```

Now we see that there is another NA, but it's not `NA` from R, instead it was loaded as character and is now considered as a category.
Well proceed here by removing that individual with that NA entry. Since this keeps an empty category for Gender, I'm also using droplevels() to get rid of it.

```{r}
d4 <- d3 %>% dplyr::filter( !(Gender %in% c("NA","N")) ) %>% droplevels()
skimr::skim(d4)
```

All done, data is clean now.

Let's assign at the end to some final variable, this makes it easier to add further cleaning steps above.

>>>>

---
title: "Cleaning HR Data for Analysis"
author: "Guillermo Gallardo"
date: "Summer 2024"
output: html_document

@@

```
```{r}
tenure_breaks <- seq(0, 40, by = 5)
tenure_labels <- paste0(tenure_breaks[-length(tenure_breaks)], "-", tenure_breaks[-1], " years")

d3 = d2 %>%
  mutate(TenureCategory = cut(YearsAtCompany, breaks = tenure_breaks, labels = tenure_labels, include.lowest = TRUE))

head(d3)
```

```{r}
HRprocesseddata = d3
```

@@

### 2024-06-15 products/manuscript/Manuscript.qmd
Project part 2 - second update

**Removing Column**

@@

**Adding Labels**

@@

@@

@@

```{r}
#| label: col-hrgender
#| tbl-cap: "Employee Gender"
#| echo: FALSE
gender_table=readRDS("../../results/tables/HRgendertable.rds")
knitr::kable(gender_table)
```

**MAYBE ADD LABEL FOR TENURE?**

### **Education** MOVE TO EDA

During data cleaning, we'll use the data dictionary to decode each field. For instance, the education field is labeled from 1 to 5, representing different educational levels: 1 = "Below College," 2 = "College," 3 = "Bachelor's," 4 = "Master's," and 5 = "Doctorate." Notably, about 40% of the dataset holds a Bachelor's degree.

```{r }

```

>>>>

**Removing Columns**

@@

**Adding Descriptive Labels**

@@

**Tenure Categorization**

To enhance the visualization of tenure distribution, I opted to introduce a new variable categorizing employee tenure. This approach allows us to identify the predominant tenure group, offering insights into potential turnover risks.

```{r, results='hide'}
tenure_breaks <- seq(0, 40, by = 5)
tenure_labels <- paste0(tenure_breaks[-length(tenure_breaks)], "-", tenure_breaks[-1], " years")

d3 = d2 %>%
  mutate(TenureCategory = cut(YearsAtCompany, breaks = tenure_breaks, labels = tenure_labels, include.lowest = TRUE))
```

@@

**Tenure category**

More than half of the employees have tenure of less than 5 years, suggesting a relatively young workforce. Shorter tenure could potentially indicate a higher turnover risk, although a comprehensive analysis incorporating other variables is necessary to fully understand the situation.

```{r}
#| label: fig-tenureCategory
#| fig-cap: "Tenure Category"
#| echo: FALSE
knitr::include_graphics(here("results","figures","tenure-category.png"))
```

@@

```{r}
#| label: tbl-hrgender
#| tbl-cap: "Employee Gender"
#| echo: FALSE
gender_table=readRDS("../../results/tables/HRgendertable.rds")
knitr::kable(gender_table)
```

### **Monthly Income and Education**

I'm interested in examining which educational group experiences higher attrition rates. It's evident from the data that median salaries increase with higher levels of education, but there are also numerous outliers present. This visualization serves to highlight the relationship between monthly income and educational background, which could provide insights into how these factors influence employee turnover.

```{r }
#| label: fig-Income-Education
#| fig-cap: "Monthly Income by Education Level"
#| echo: FALSE
knitr::include_graphics(here("results","figures","income-education.png"))
```
